---
Title: "LA Mortality Tutorial 2"
author: "Aidan Clements"
date: February 20, 2024
output: pdf_document
---
Loading readr and diplyr
```{r}
library(readr)
library(dplyr)
```

I am uploading load the LA Mortality File into RStudio 
```{r}
la_mort <- read_csv("https://www.dropbox.com/scl/fi/fzsnhfd3lq80v2o3sag6c/la_mort.csv?rlkey=h1vyjm2b8ppgejgsg3e8evm7i&dl=1")
```

LA Mortality File Codebook: 

###*restatus*, resident status

- 1= state and county of occurrence and residence are the same
- 2= state of occurrence and residence are the same, but county is different
- 3= state of occurrence and residence are different, but both in U.S.
- 4= state of occurrence is U.S. but residence is outside U.S.

### *stocr*, state of occurrence
- Abbreviation for state where death occurred

### *cntyocr*, county of occcurrence
- Abrreviation for county where death occurred

### *popcntyocr*, population of county of occurrence
Population Size of County of Occurrence
Based on the results of the 2000 Census
- 0 ... County of 1,000,000 or more
- 1 ... County of 500,000 to 1,000,000
- 2 ... County of 250,000 to 500,000
- 3 ... County of 100,000 to 250,000 9
- 9 ... County of less than 100,000 

### *strsd*, state of residence
- Abbreviation for state of residence of individual death 

### *cntyrsd*, country of residence 
- Abbreviation for county of residence of individual death 

### *popcntyrsd*, population size of country of residence
Population Size of County of Residence
Based on the results of the 2000 Census
- 0 ... County of 1,000,000 or more
- 1 ... County of 500,000 to 1,000,000
- 2 ... County of 250,000 to 500,000
- 3 ... County of 100,000 to 250,000
- 9 ... County of less than100,000
- Z ... Foreign residents 

### *stbrth*, State of birth

### *birthr*, state of birth recode

### *educ1989*, education level (1989 revision)
Education (1989 revision)
- 00: no formal education
- 01-08: Years of elementary school
- 91: year of high school
- 10: 2 years of high school
- 11: 3 years of high school
- 12 4 years of high school
- 13: 1 year of college
- 14: 2 years of college
- 15: 3 years of college
- 16: 4 years of college
- 17: 5 or more years of college
- 99: not stated

### *educ2003*, education level (2003 revision)
- 1: 8th grade or less
- 2: 9 - 12th grade, no diploma
- 3 high school graduate or GED completed
- 4: some college credit, but no degree
- 5: Associate degree
- 6: Bachelor’s degree
- 7: Master’s degree
- 8: Doctorate or professional degree
- 9: Unknown

### *educflag*, Education reporting flag
- 0 1989 revision of education item on certificate
- 1 2003 revision of education item on certificate
- 2 no education item on certificate 

### *monthdth*, Month of Death
-1: January
-2: February etc.

### *sex*, Sex
- sex of individual who died

### *age*, Age at death
Four positions are used to code detail age. Location 70 identifies age in years, months, days, etc. Locations 71-73 are the number of years, months, days, etc.
- 1: Years. 001-135,999
- 2: Months 001-011,999
- 4: Days 001-027,999
- 5: Hours 001-023,999
- 6: Minutes 001-059,999
- 9: Age not stated 999

### *placdth*, Place of death and decendent's status
Hospital, clinic or Medical Center
- 1: Inpatient
- 2: Hospital, Clinic or Medical Center- Outpatient or admitted to Emergency Room 
- 3: Hospital, Clinic or Medical Center- Dead on Arrival
- 4: Decedent’s home
- 5:  Hospice facility
- 6: Nursing home/long term care
- 7:  Other
- 9: Place of death unknown

### *marstat*, Marital Status
- S: Never married, single
- M: Married
- W: Widowed
- D: Divorced
- U: Marital Status unknown

### *weekday*, Day of Week of Death
- 1:Sunday
- 2: Monday
- 3: Tuesday
- 4: Wednesday
- 5: Thursday
- 6: Friday
- 7: Saturday
- 9: Unknown

### *year*, Current Data Year

### *mandeath*, Manner of Death
- 1: Accident
- 2: Suicide
- 3: Homicide
- 4: Pending investigation
- 5: could not determine
- 6: self-inflicted
- 7: natural
- Blank: Not Specified

### *prgnt*, Pregnancy Tag

### *ucod*, Underlying Cause of Death (ICD code 10)

### *ucr358*, 358 cause recode

### *ucr113*, 113 cause recode

### *race*, Bridged Race
- 01: White
- 02: Black
- 03: American Indian
- 04: Chinese
- 05: Japanese
- 06: Hawaiian
- 07: Filipino
- 18: Asian indian
- 28: Korean
- 38: Samoan
- 48: Vietnamese
- 58: Guamanian
- 68: Other asian or pacific islander
- 78: combined other asian or pacific islander

### *racer3*, race recode 3
- 1: white
- 2: races other than white or black
- 3: Black

### *racer5*, race recode 5
- 0: Other (Puerto Rican Only)
- 1: White
- 2: Black
- 3: American Indian
- 4: Asian or Pacific Islander

### *hispanic*, hispanic origin
- 100-199: Non-Hispanic
- 200-209: Spaniard
- 210-219: Mexican
- 260-269: Puerto Rican
- 270-274: Cuban
- 275-279: Dominican
- 220: Central and South American
- 221-230: Central American
- 231-249: South American
- 250-259: Latin American
- 280-299: Other Hispanic
- 996-999: Unknown










